Shallow EDSLs and Object-Oriented Programming

### 2016-06-28 paper/sections/Conclusion.tex
Conclusion
4 pages

One limitation of the techniques presented here is that
transformations require code duplication in extensions.  For
transformations, we have to refine their return type in extensions and
supply a new implementation by duplicating the original code. 
Though these methods can still be defined modularly with advanced type
system features of Scala~\cite{zenger05independentlyextensible},
the encoding would become significantly more complicated.
How to define these transformations modularly, without code
duplication, and without sophisticated types
is a possible line of future work.  Also, it is tedious to
express the inheritance relationships in extensions, especially when
multiple inheritance is used. Another line of future work is to use
some meta-programming mechanisms to eliminate such boilerplate.

>>>>

One limitation of the techniques presented here is that
transformations require code duplication in extensions.
We have to refine their return type and
supply a new implementation by copying the original code.
Though transformations can still be defined modularly with advanced type
system features of Scala~\cite{zenger05independentlyextensible},
the encoding would become significantly more complicated.
How to define these transformations modularly, without code
duplication, and without sophisticated types
is a possible line of future work.  Also, it is tedious to
express the inheritance relationships in extensions, especially when
multiple inheritance is used. Another line of future work is to use
some meta-programming mechanisms to eliminate such boilerplate.

### 2016-06-28 paper/sections/Introduction.tex
4 pages

In this paper, to avoid ambiguity, we prefer a more precise meaning
for shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises from the domain of shallow DSLs being
typically a function, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of Object-Oriented
Programming. Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. 
%This should
%not came as a surprise, as OOP languages have evolved over more than 50
%years to improve the use of procedural abstraction. 
In order to understand how OOP can help with shallow embeddings, it is
helpful to
first review some of the limitations commonly
found in the literature:
\vspace{-5pt}
\begin{enumerate}

@@

\end{enumerate}
\vspace{-5pt}

@@

We show that OOP abstractions, including \emph{
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. From the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.

>>>>

In this paper, to avoid ambiguity, we prefer a more precise meaning
for shallow embeddings as EDSLs implemented using \emph{procedural abstraction}~\cite{reynolds94proceduralabstraction}. Such
interpretation arises from the domain of shallow EDSLs being
typically a function, and procedural abstraction being a way to encode
data abstractions using functions. As Cook~\cite{cook09abstraction} argued,
procedural abstraction is also the essence of Object-Oriented
Programming. Thus, according to our definition, the implementation of a shallow
EDSL in OOP languages should simply correspond to a standard
object-oriented program.

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. 
%This should
%not came as a surprise, as OOP languages have evolved over more than 50
%years to improve the use of procedural abstraction. 
In order to understand how OOP can help with shallow embeddings, it is
helpful to
first review some of the limitations commonly
found in the literature:
\begin{enumerate}

@@

\end{enumerate}

@@

We show that OOP abstractions, including \emph{
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanisms enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala\footnote{Available online: \url{https://github.com/wxzh/shallow-dsl}}.
From the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.

### 2016-06-28 paper/sections/Transformations.tex
4 pages

\paragraph{Desugaring in Scala}
A corresponding Scala implementation is given below:
\lstinputlisting[linerange=4-17]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations.
Implementing it for each case is
straightforward: \emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} just mimics the
formula. Circuits are constructed using the helper methods:
\lstinputlisting[linerange=3-6]{../src/desugar/package.scala}%APPLY:COMPANION

>>>>

\paragraph{Desugaring in Scala}
A corresponding Scala implementation is given below:
\lstinputlisting[linerange=4-17]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations.
Implementing each case is straightforward:
\emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} mimics the
formula.

Circuits are constructed via some helper methods:
\lstinputlisting[linerange=3-6]{../src/desugar/package.scala}%APPLY:COMPANION
These methods would be intensively used in client code, e.g.
\lstinputlisting[linerange=5-6]{../src/test/Test.scala}%APPLY:CLIENT
By importing a different package, clients can obtain new semantics with the code
describing circuits untouched.
